---
title: "Quito PM2.5 data files tidying 2nd part"
author: "Alain Rodriguez"
date: "25 de octubre de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Loading libraries
```{r loading_libraries, message=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
library(dplyr)
```

We go now for the **2008 - 2016** period file, as it has a very different format from that of previous years.

## 2. Loading data files
```{r loading_data, warning=FALSE}
pm25_2008_2016<- read_xlsx('2_PM25_2008_2016.xlsx', col_names = TRUE, skip = 1)
head(pm25_2008_2016, 5)
```

It appears as a very messy file with observations as variables and viceversa. Let's try to tidy this a bit
```{r}
pm25_2008_2016<- pm25_2008_2016 %>% mutate(FECHA = paste(`AÑO`, MES, DIA, sep = '-')) %>%
        select(-c(MAGNITUD, `AÑO`, MES, DIA)) %>% gather(key = HORA, value = DATO, -c(ESTACION, FECHA))
pm25_2008_2016
        
```
We need to perform a few operations with the **FECHA** and **HORA** columns. Firstly, we're going to recode the **HORA** column as 'HORA 1' = 01:00:00, 'HORA 2' = 02:00:00 and so on till 'HORA 24' = 24:00:00
We do this using gsub with regular expressions.
```{r}
pm25_2008_2016$HORA<- gsub("(HORA) (\\d+)", "\\2:00:00", pm25_2008_2016$HORA)
```
After that let's *unite* columns **FECHA** and **HORA** and *spread* the observations across the **ESTACION** columns.
```{r}
pm25_2008_2016 %<>% unite(FECHA, FECHA, HORA, sep = " ") %>% spread(key = ESTACION, value = DATO)
```
Finally we turn **FECHA** column into a real date-time column
```{r}
pm25_2008_2016$FECHA<- parse_date_time(pm25_2008_2016$FECHA, orders = 'ymd HMS')
```
Let's take a look at the results
```{r}
pm25_2008_2016 %>% arrange(FECHA)
```
```{r plotting_2, echo = FALSE}
pm25_2008_2016 %>% filter(FECHA > as.Date('2008-01-01') & FECHA < as.Date('2008-01-31')) %>% ggplot(aes(FECHA, CENTRO)) + geom_col() + scale_x_datetime(date_labels = '%b-%Y')
```
```

